### 2022-10-05 AISTATS2023/paper.tex
updated real example part

The simulation examples, where the data-generating distribution
\(p(\mathbf{X})\) and the predictive function \(f(\cdot)\) are defined
by us, enable the evaluation of competitive approaches against a solid
ground-truth. We follow this common XAI
practice~\citep{aas2021explaining, herbinger2022repid} for providing
secure empirical evidence about the superiority of UALE method.

We split the simulation examples in two groups. The first group,
Section~\ref{sec:simulation-examples-1}, aims at showcasing that UALE
method is more accurate than PDP-ICE in quantifying the average effect
and the level of heterogeneous effects (uncertainty), when input
features are correlated. The second group,
Section~\ref{sec:simulation-examples-2}, illustrates that UALE method
achieves a better approximation (average effect and uncertainty) in
cases of limited samples, due to automatic bin-splitting. In both
groups, we choose to illustrate the most indicative examples; a more
extensive evaluation is provided in the Appendix.

@@

In this simulation, we will compare UALE method against PDP-ICE in
quantifying the main effect and the uncertainty. Since there is
ambiguity about the ground-truth first-order effect in cases of
correlated features,
e.g.~\citep{apley2020visualizing,Gromping2020MAEP}, we limit ourselves
to the following piecewise linear function,

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $f_{\mathtt{lin}} < 0.5$ }\\
                    0.5 - f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $0.5 \leq f_{\mathtt{lin}} < 1$}\\
                    \alpha f_{\mathtt{int}} &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_{int}(\mathbf{x}) = x_1x_3\). As we observe, the linear term
\(f_{\mathtt{lin}}\) includes the two correlated features and the term
\(f_{\mathtt{int}}\) interacts the two non-correlated variables. The
samples that we use in all examples are coming from the following
distribution: \(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We will test the effect
computed by UALE and PDP-ICE in three cases; (a) no interaction
(\(\alpha=0\)) and equal weights (\(a_1=a_2\)), (b) no interaction
(\(\alpha=0\)) and different weigths (\( a_1 \neq a_2 \)) and (c) with
interaction (\(\alpha \neq 0\)) with equal weights (\(a_1=a_2\)).

In all cases, we firstly compute the ground-truth average effect and
the uncertainty analytically (proofs in the Appendix) and then we
compare it against the approximation provided by each method. The
approximation is computed after generating \(N=300\) samples. As we
will see, despite the model's simplicity, PDP-ICE fail in quantifying
the effect correctly, due to correlation between features \(X_1\) and
\(X_2\). Finally, besides it is not the focus of the current examples,
we will also observe that UALE manages to correctly extract the
regions with constant effect. Details for obtaining the ground truth
effects and the experimental set-up are in the Appendix.

@@

We test the feature effect when no interaction term is apparent,
i.e. \(\alpha=0\), and the weights are equal \(a_1=a_2=1\).  In this
case, the ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(x_1\)
when \(0 \leq x_1 < 0.25\), \(-x_1\) when \(0.25 \leq x_1 < 0.5\) and
zero otherwise. In each position, there is total absence of
heterogeneous effects; therefore, the uncertainty of the global
explanation is zero \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP main effect is
wrong and ICE plots imply the existence of heterogeneous effects. In
contrast, ALE captures correctly the average effect and the zero
uncertainty and it also groups perfectly the regions with
constant-effect.

@@

As before, there is no interaction term and, therefore, the
ground-truth is zero, i.e.  \(f^{\mathcal{GT}}_{\sigma^2}(x_1) =
0\). The weigts are different \(a_1=2, a_2=0.5\), therefore, the
ground-truth effect is \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(2x_1\) when
\(0 \leq x_1 < 0.2\), \(-2x_1\) when \(0.2 \leq x_1 < 0.4\) and zero
otherwise. In Figure~\ref{fig:ex-synth-1-2}, we observe that PDP
estimation is completely opposite to the ground-truth effect,
i.e.~negative in \([0, 0.2)\) and positive in \([0.2, 0.4)\), and the
ICE erroneously implies heterogeneous effects.  As before, ALE
quantifies perfectly the ground truth effect and the zero-uncertainty,
extracting correcltly the constant effect regions.

@@

In this case we activate the interaction term, i.e. \(a=1\), and we
set equal weights \(a_1=a_2=1\). The interaction term provokes
heterogeneous effects in features \(x_1, x_3\), i.e., at any position
\(x_1\), the local effects dependend on the unknown value of \(X_3\)
and vice-versa. The effect of \(x_2\) continues to have
zero-uncertainty, since it does not appear in any interaction term. As
we observe in Figure~\ref{fig:ex-synth-1-3}, UALE captures perfectly
the effect of all features, and the uncertainty in all cases. As
expected, PDP-ICE quantify the effect and the uncertainty correctly
only in the case of \(x_3\), since \(X_3\) is independent from other
features. For the correlated features, i.e. \(x_1, x_2\), the average
effect computed by PDP and the uncertainty implied by ICE plots are
wrong.

@@

% \paragraph{Discussion.}

% In purpose, the examples above do not cover the case where there is an
% interaction term includes correlated features, i.e. \(x_1x_2\). In
% this scenario, there is an open debate about the ground-truth
% effect~\citep{Gromping2020MAEP}.

@@

In this simulation, we aim to to quantify the advantages of automatic
bin-splitting based on the objective of Eq.(\ref{eq:opt}), through the
following validation framework. We generate a big dataset with dense
sampling (\(N=10^6\)) and we treat the DALE estimation with dense
fixed-size bins (\(K=10^3\)) as ground-truth. Afterwards, we generate
less samples (\(N=500\)) and we compare (a) the fixed-size DALE
estimation for many different \(K\) versus (b) the auto-bin splitting
algorithm. In all cases, we use a bivariate black-box function
\(f(\cdot)\) where the samples are instances of the distribution
\(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\) and
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\).

We denote as
\(\mathcal{Z^{\mathtt{DP}}} = \{z^{\mathtt{DP}}_{k-1}, \cdots,
z^{\mathtt{DP}}_{K}\}\) the sequence obtained by automatic
bin-splitting based on the optimisation problem of Eq.~(\ref{eq:opt})
and with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size splitting with
\(K\) bins. The evaluation is done with regard to two metrics;
\(\mathcal{L}_{\mathtt{DP|K}}^{\mu} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} | \mu_k - \hat{\mu}_k | \) quantifies
the average error in estimating the the expected effect and
\(\mathcal{L}_{\mathtt{DP|K}}^{\sigma} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} | \sigma_k - \hat{\sigma}_k | \)
the average error in estimating the variance of the effect. The metric
\(\mathcal{L}^{\rho}_{\mathtt{DP|K}} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} \rho_k \) quantifies the average
nuisance uncertainty. The optimal bin-splitting is the one that
minimises, at the same time, both \(\mathcal{L}^{\mu}\) and
\(\mathcal{L}^{\sigma^2}\) errors. For consistent results, in all the
examples below, we regenerate samples and repeat the computations for
\(t = 10\) times, providing the mean value for all metrics.

@@

The California Housing is a largely-studied dataset with approximately
\(20000\) training instances, making it appropriate for robust
Monte-Carlo approximations. The dataset contains \(D=8\) features
describing building blocks of California, like latitude, longitude,
total number of people residing in the block etc., and the target is
the median house value inside the block. We exclude instances with
missing or outlier values, i.e. \(x_d^i > \hat{sigma}_j\) in any
feature and normalize them to have
\(\hat{\mu}_d = 0, \hat{\sigma_d}\). We split the dataset into
\(N_{tr} = 15639\) training and \(N_test = 3910\) test examples (80/20
split) and we fit a Neural Network with 3 hidden layers of 256, 128
and 36 units respectively. After 15 epochs using the Adam optimizer
with learning rate \(\eta = 0.02\), the model achieves a normalized
mean squared error (R-Squared) of \(0.25\) (0.75). In depth information
about preprocessing, training and evaluation are provided in the
Appendix.

@@

In real-world datasets, it is infeasible to obtain the ground truth FE
for seamlessly evaluating the competitive methods. Therefore, the main
purpose of this part is to demonstrate the usefulness of uncertainty
quantification in UALE, i.e., illustrate that modeling the
heterogeneous effects is important for the interpretation of an ALE
plot. For this reason, Secondly, we discuss about the FE plots obtaing with UALE and
PDP-ICE.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_1_ale_auto.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_1_pdp_ice.pdf}\\
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_7_ale_auto.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_7_pdp_ice.pdf}\\
  \caption{Figure 1}
  \label{fig:ex-synth-2-1}
\end{figure}

@@

For evaluating UALE in terms of bin-splitting we set-up the following
evaluation framework. We treat as ground-truth the effects computed on
the full training-set \(N=20000\) with dense fixed-size bin-splitting
(\(K=80\)). Given the big number of samples, we make the hypothesis
that the approximation with dense binning is close to the ground
truth. Afterwards we select less samples \(N=1000\) and we compare
UALE approximation with all possible fixed-size alternatives, in terms
of accuracy, i.e. we provide the quantities
\(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma^2}, \mathcal{L}^{\rho^2}\).
\section{CONCLUSION}

>>>>

Simulation examples provide the freedom to select the data-generating
distribution \(p(\mathbf{X})\) and the predictive function
\(f(\cdot)\). We follow this common XAI
practice~\citep{aas2021explaining, herbinger2022repid} for empirically
evaluating UALE against a solid ground-truth. The first group of
synthetic examples (Section~\ref{sec:simulation-examples-1}) compares
UALE with PDP-ICE plots in quantifying the average and the uncertainy
of the effect. The second group
(Section~\ref{sec:simulation-examples-2}) evaluates the accuracy of
the approximation (average effect and uncertainty) using the automatic
bin-splitting algorithm. In both groups, we illustrate the most
indicative examples; a more extensive evaluation is provided in the
Appendix.

@@

In this simulation, we will compare UALE method against PDP-ICE in
quantifying the main effect and the uncertainty. The samples that we
use in all examples are coming from the following distribution:
\(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We create the following
piecewise linear function:

% Since there is
% ambiguity about the ground-truth first-order effect in cases of
% correlated features,
% e.g.~\citep{apley2020visualizing,Gromping2020MAEP},

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $f_{\mathtt{lin}} < 0.5$ }\\
                    0.5 - f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $0.5 \leq f_{\mathtt{lin}} < 1$}\\
                    \alpha f_{\mathtt{int}} &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_{\mathtt{int}}(\mathbf{x}) = x_1x_3\). The linear term
\(f_{\mathtt{lin}}\) includes the two correlated features and
\(f_{\mathtt{int}}\) interacts the two non-correlated features.  We
evaluate the effect computed by UALE and PDP-ICE in three cases; (a)
without interaction (\(\alpha=0\)) and equal weights (\(a_1=a_2\)),
(b) without interaction (\(\alpha=0\)) and different weigths
(\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha \neq 0\)) with
equal weights (\(a_1=a_2\)).

In all cases, we provide the ground-truth average effect and
uncertainty computed with analytical derivations (proofs at the
Appendix) and we compare it against the approximation provided by each
method. The approximation is computed after generating \(N=500\)
samples.

@@

We test the feature effect when no interaction exits \(\alpha=0\) and
the weights are equal \(a_1=a_2=1\). The ground truth effect
\(f_\mu^{\mathtt{GT}}(x_1)\) is: \(x_1\) when \(0 \leq x_1 < 0.25\),
\(-x_1\) when \(0.25 \leq x_1 < 0.5\) and zero otherwise. There zero
heterogeneous effects, therefore
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP main effect is
wrong and ICE plots imply the existence of heterogeneous effects. In
contrast, ALE captures correctly the average effect and the zero
uncertainty and it also groups perfectly the constant-effect regions.

@@

As before, no interaction exits \(\alpha=0\), therefore, the
ground-truth uncertainty is zero, i.e.
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). The weigts are
\(a_1=2, a_2=0.5\), therefore, the ground-truth effect is
\(f_\mu^{\mathtt{GT}}(x_1)\) is, \(2x_1\) when \(0 \leq x_1 < 0.2\),
\(-2x_1\) when \(0.2 \leq x_1 < 0.4\) and zero otherwise. In
Figure~\ref{fig:ex-synth-1-2}, we observe that PDP estimation is
completely opposite to the ground-truth effect, i.e.~negative in the
region \([0, 0.2)\) and positive in \([0.2, 0.4)\), and the ICE
erroneously implies the existence of heterogeneous effects. As before,
ALE quantifies correctly the ground truth effect, the
zero-uncertainty and extracts the constant effect regions.

@@

In this case we activate the interaction term, i.e. \(a=1\), and we
set the weights to \(a_1=a_2=1\). The interaction term provokes
heterogeneous effects for features \(x_1, x_3\), because the local
effects at \(x_1\) depend on the unknown value of \(X_3\) and
vice-versa. The effect of \(x_2\) has zero-uncertainty since it does
not appear in any interaction term. As we observe in
Figure~\ref{fig:ex-synth-1-3}, UALE correcly models the average effect
and the uncertainty in all cases. PDP-ICE quantifies correctly the
effect and the uncertainty only in the case of feature \(x_3\),
because \(X_3\) is independent from other features. For features
\(x_1, x_2\) that are correlated the average effect of PDP
and the uncertainty of ICE plots are wrong.

@@

\paragraph{Discussion.}

Despite the model's simplicity, PDP-ICE fail in quantifying the
average effect and the uncertainty due to correlation between features
\(X_1\) and \(X_2\). UALE quantifies them correctly and estimates them
accurately extracting the regions with constant effect. The examples
above do not cover the case of an interaction term between correlated
features, e.g. \(x_1x_2\), because there is an open debate about the
ground-truth effect in this case~\citep{Gromping2020MAEP}.

@@

In this simulation, we aim to to quantify the advantages of automatic
bin-splitting. We set-up the following validation framework. We
generate a big dataset with dense sampling (\(N=10^6\)) and we treat
the DALE estimation with dense fixed-size bins (\(K=10^3\)) as
ground-truth. Afterwards, we generate less samples (\(N=500\)) and we
compare the fixed-size DALE estimation (for many different \(K\))
versus the automatic bin-splitting. In all cases, we use a bivariate
black-box function \(f(\cdot)\) where the samples are instances of the
distribution \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\) and
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We denote as
\(\mathcal{Z^{\mathtt{DP}}} = \{z^{\mathtt{DP}}_{k-1}, \cdots,
z^{\mathtt{DP}}_{K}\}\) the sequence obtained by automatic
bin-splitting and with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size
splitting with \(K\) bins.  For consistent results, in all the
examples below, we regenerate samples and repeat the computations for
\(t = 30\) times, providing the mean value for all metrics.

\paragraph{Metrics}

The evaluation is done with regard to two metrics that count the mean
bin error, where bin error is the absolute difference of the
approximation from the ground-truth. The metric
\(\mathcal{L}_{\mathtt{DP|K}}^{\mu} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} | \mu_k - \hat{\mu}_k | \) counts the
mean bin error on the average effect and
\(\mathcal{L}_{\mathtt{DP|K}}^{\sigma} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} | \sigma_k - \hat{\sigma}_k | \) the
mean bin error on the uncertainty. In both cases the ground truth is
computed from the dense bins of ground-truth that overlapse with the
wider bin of the approximation. We also provide the mean (per bin)
error term
\(\mathcal{L}^{\rho}_{\mathtt{DP|K}} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} \rho_k \) for a complete understanding
of the uncertainty error.

@@

The California Housing is a largely-studied dataset with approximately
\(20000\) training instances, making it appropriate for robust
Monte-Carlo approximations. The dataset contains \(D=8\) features
describing building blocks of California, like latitude, longitude,
total number of people residing in the block etc., and the target is
the median house value inside the block. We exclude instances with
missing or outlier values, i.e. \(x_d^i > \hat{\sigma}_j\) in any
feature and normalize them to have
\(\hat{\mu}_d = 0, \hat{\sigma_d}\). We split the dataset into
\(N_{tr} = 15639\) training and \(N_{test} = 3910\) test examples (80/20
split) and we fit a Neural Network with 3 hidden layers of 256, 128
and 36 units respectively. After 15 epochs using the Adam optimizer
with learning rate \(\eta = 0.02\), the model achieves a normalized
mean squared error (R-Squared) of \(0.25\) (0.75). In depth
information about preprocessing, training and evaluation are provided
in the Appendix.

@@

In real-world datasets, it is infeasible to obtain the ground truth FE
for seamlessly evaluating the competitive methods. For this reason, we
selected the California Dataset, where, in broad-terms, UALE and
PDP-ICE aggree in the quantification of the average effect and of the
uncertainty. In Figure\ref{fig:ex-real-1}, we observe that UALE
complements ALE for the quantification of the heterogeneous effects,
similarly to as ICE complements PDP. Add some content..

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_1_ale_auto.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_1_pdp_ice.pdf}\\
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_4_ale_auto.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_4_pdp_ice.pdf}\\
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_5_ale_auto.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_5_pdp_ice.pdf}\\
  \caption{Figure 1}
  \label{fig:ex-real-1}
\end{figure}

@@

As in the Syntetic Example (Case 2), we treat as ground-truth the
effects computed on the full training-set \(N=20000\) with dense
fixed-size bin-splitting (\(K=80\)). Given the big number of samples,
we make the hypothesis that the approximation with dense binning is
close to the ground truth. Afterwards, we randomly select \(N=1000\)
samples and we compare UALE approximation with all possible fixed-size
alternatives using \(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma}\). Add discussion

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_1.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_1.pdf}\\
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_4.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_4.pdf}\\
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_5.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_5.pdf}\\
  \caption{Figure 1}
  \label{fig:ex-real-2}
\end{figure}

\paragraph{Discussion}

\section{CONCLUSION}
